Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2021-12-18 analysis/paper/paper.Rmd
most comments addressed

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the {glmmLDTS} package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior.

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such,
results are shown for all ages. Bearded seals were predicted to have a bi-modal
distribution of haul-out probability across the day (Figure
\@ref(fig:beardedHOCal)). In addition to a peak around local solar noon, the
bearded seal model predicts additional haul-out activity around local midnight.
The haul-out behavior is also more protracted throughout the spring season
compared to ribbon and spotted seals.

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that is less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
Adults of both sexes spend considerable time in April, May and June resting on the
ice. In contrast to ribibon seals, adult spotted seal males have a more protracted
haul-out season compared to females, and more time spent resting in June and
July (Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with 
ribbon seals, the young-of-the-year records begin after weaning and
the model predictions demonstrate the ontogeny of in-water activities (e.g.
diving, foraging) in May.

@@

We modeled data from bio-loggers deployed on bearded, ribbon, and spotted seals
to examine factors affecting haul-out behavior on sea ice in the Bering and
Chukchi seas. Our analysis shows seals haul out progressively more through the
spring and peak near mid-May to early June before declining again. This pattern
aligns well with what has been previously documented and confirms our haul-out
data are likely capturing population-level behavioral patterns. Seals
preferentially haul out on ice shortly after solar noon, which allows seals to
maximize absorption of solar radiation thought to facilitate the molting process
[@feltz1966]. Interestingly, bearded seals appear to have two peaks in haul-out
activity within a day, one shortly after solar noon, and one centered near solar
midnight. A similar bi-modal pattern has been seen in ringed seals
[@vonduyke2020] and suggests that bearded and ringed seals may be operating
under different constraints than ribbon and spotted seals. Bearded and ringed 
seals are distributed across higher latitudes and the extended daylight hours
may allow more flexibility in optimizing resting periods with foraging. Other
factors such as polar bear predation (which is rare for ribbon and spotted seals
in the Bering Sea) may also explain differing haul-out patterns. Bearded seals also
showed a more protracted season of haul-out behavior compared to ribbon and
spotted seals. This aligns with findings from Thometz et al. [@thometz2021] who
observed a mean molting period of 119$\pm$2 days and a relatively stable resting
metabolic rate during that time. In the same study, spotted and ringed seals
underwent molt periods of 33$\pm$4 and 28$\pm$6 days and exhibited increased
resting metabolic rates.

Unlike previous analyses of seal haul-out data (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Ribbon seals are not known to form triads and our model
predicts a progression of increased haul-out behavior with females starting
earlier in the season than males. Notably, female ribbon seals spend a large
portion of the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. Spotted
seals are known to form triads during the breeding season [@burns2009,
@mcclintock2015] where a female and dependent pup are accompanied on the ice by
a suitor male. The male waits for the female to wean the pup and enter estrus,
and fends off any other males that may be interested. Triad formation results in
both males and females spending a large portion of the day hauled out on ice and
a protracted spring haul-out season for both sexes. Because mating occurs in the
water, females may also be less inclined to venture out for foraging trips while
still nursing the pup. We see this reflected in the predicted haul-out behavior,
with both males and females exhibiting a broad distribution of time out of the
water throughout the solar day and the season.  In the case of both ribbon and
spotted seals, subadults are the first to begin consistent haul-out behavior and
follow a typical phocid pattern where subadults molt first as they do not have
any reproductive constraints.

@@

We recognize that the species and ecosystems studied in this paper are
within the traditional range of the Inpuiaq and Yupik people. And, we
would like to acknowledge and thank them as the original stewards of
these ecosystems. The deployment of bio-logging devices used in this
study were often done in collaboration with Alaska Native seal hungers and their communities.
We would like to especially acknowledge the communities of Kotzebue,
Utqiaġvik, and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris,
Grover Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin,
Henry Goodwin, John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk,
Noah Naylor, Virgil Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen
Stone, and Randy Toshavik from Kotzebue, Alaska; Billy Adams, Mary-Ellen
Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard
Kittick, Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik,
Shawn Oktollik, Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross
Schaeffer, Bob Shears, and Joe Skin from the North Slope Borough,
Alaska;

We are grateful for the assistance in catching and sampling seals by
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir
Burkanov, Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey
Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson,
Charles Littnan, Lloyd Lowry, Erin Moreland, Lorrie Rea, Gay Sheffield,
Brent Stewart, and Dave Withrow. We also appreciate the commitment to
science and safety by all officers and crew of the NOAA Ship Oscar
Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.

@@

The findings and conclusions in the paper are those of the author(s) and do not
necessarily represent the views of the National Marine Fisheries Service, NOAA.
Any use of trade, product, or firm names does not imply an endorsement by the
U.S. Government. Funding for this study was provided by the U.S. National
Oceanic and Atmospheric Administration. The field work was conducted under the
authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
Funding to ADFG for tagging seals was provided by the Bureau of Ocean Energy
Management (No. M13PC0015) and the Office of Naval Research (No.
N00014-16-1-3019). ADFG and NSB field work was covered by Research Permits Nos.
358-1585, 358-1787, 15324, and 20466 and by ADFG IACUC permits Nos. 06-16,
09-21, 2010-13, 2012-020, 0027-2019-41, 2013-020, 2014-03,2015-25, 2016-23,
0027-2017-27, 0027-2018-29, 0027-2019-041.

>>>>

To properly account for temporal dependence within a computationally tractable
framework, we used generalized linear mixed pseudo-models (GLMPMs;
[@verhoef2010a]) to model variation in haul-out behavior as a function of (1)
covariate predictors, (2) temporally autocorrelated random effects, and (3)
individual random effects representing heterogeneity in individual behavior. We
used the {glmmLDTS} package [@verhoef2010a] to implement GLMPMs. We explored two
different model formulations for our data, and owing to the large number of
records, we fit separate models to bearded, ribbon, and spotted seal data sets.
In our first model formulation, for each species, we fitted a year-independent
model that predicted average haul-out behavior as a function of demographic,
environmental, seasonal, and diurnal effects. Second, for ribbon and spotted
seals (which had considerably more data than bearded seals), we fitted models
that included all the effects from the first model, but also permitted annual
variation in haul-out timing. This second class of models was used to examine
whether haul-out patterns varied by year and to determine the annual timing of
apparent peaks in haul-out behavior. For both models, the response variable was
the transmitted hourly percent-dry value.

@@

Age and sex class were not included in the model for bearded seals due to our
lower sample size for adult and young-of-year age classes. As such,
results are shown for all ages. Bearded seals had a bi-modal
distribution of haul-out probability across the day (Figure
\@ref(fig:beardedHOCal)). In addition to a peak around local solar noon, the
bearded seal model predicts additional haul-out activity around local midnight.
The haul-out behavior is also more protracted throughout the spring season
compared to ribbon and spotted seals.

@@

Compared to ribbon seals, spotted seals showed a longer spring haul-out season
that is less intensely centered on solar noon (Figure \@ref(fig:spottedHOCal)).
Adults of both sexes spend considerable time in April, May and June hauled out.
In contrast to ribibon seals, adult spotted seal males have a more protracted
haul-out season compared to females, and more time out of the water in June and
July (Figures \@ref(fig:ribbonHOCal) and \@ref(fig:spottedHOCal)). As with
ribbon seals, the young-of-the-year records begin after weaning and the model
predictions demonstrate the ontogeny of in-water activities (e.g. diving,
foraging) in May.

@@

We modeled data from bio-loggers deployed on bearded, ribbon, and spotted seals
to examine factors affecting haul-out behavior on sea ice in the Bering and
Chukchi seas. Our analysis shows all three species of seal haul out
progressively more through the spring and peak near mid-May to early June before
declining again. This pattern aligns well with what has been previously
documented and confirms our haul-out data are likely capturing population-level
behavioral patterns. Seals preferentially haul out on ice shortly after solar
noon, which allows seals to maximize absorption of solar radiation thought to
facilitate the molting process [@feltz1966]. Interestingly, bearded seals appear
to have two peaks in haul-out activity within a day, one shortly after solar
noon, and one centered near solar midnight. A similar bi-modal pattern has been
seen in ringed seals [@vonduyke2020] and suggests that bearded and ringed seals
may be operating under different constraints than ribbon and spotted seals.
Bearded and ringed seals are distributed across higher latitudes and the
extended daylight hours may allow more flexibility in optimizing resting periods
with foraging. Other factors such as polar bear predation (which is rare for
ribbon and spotted seals in the Bering Sea) may also explain differing haul-out
patterns. Bearded seals also showed a more protracted season of haul-out
behavior compared to ribbon and spotted seals. This aligns with findings from
Thometz et al. [@thometz2021] who observed a mean molting period of 119$\pm$2
days and a relatively stable resting metabolic rate during that time. While
ribbon seals were not considered in the same study, spotted and ringed seals
underwent molt periods of 33$\pm$4 and 28$\pm$6 days and exhibited increased
resting metabolic rates.

Unlike previous analyses of seal haul-out behavior (e.g. [@verhoef2010a],
[@conn2014a]), we also investigated the influence of sex-age class on haul-out
probabilities for all species except bearded seals because of low sample size.
While both ribbon and spotted seals exhibited a unimodal diel haul-out pattern
generally centered around local solar noon, there were key differences across
species, age, and sex that match expectations given what we know of their
ecological behavior. Ribbon seals are not known to form triads and our model
predicts a progression of increased haul-out behavior with females starting
earlier in the season than males. Notably, female ribbon seals spend a large
portion of the day in the water during the pupping period, aligning with the
hypothesis that ribbon seal females continue foraging while nursing. Spotted
seals are known to form triads during the breeding season [@burns2009,
@mcclintock2015] where a female and dependent pup are accompanied on the ice by
a suitor male. The male waits for the female to wean the pup and enter estrus,
and fends off any other potential suitor males. Triad formation results in both
males and females spending a large portion of the day hauled out on ice and a
protracted spring haul-out season for both sexes. Because mating occurs in the
water, females may also be less inclined to interrupt their haul-out with
foraging trips while still nursing the pup. We see this reflected in the
predicted haul-out behavior, with both males and females exhibiting a broad
distribution of time out of the water throughout the solar day and the season.
In the case of both ribbon and spotted seals, subadults are the first to begin
consistent haul-out behavior and follow a typical phocid pattern where subadults
molt first as they do not have any reproductive constraints.

@@

We recognize that the species and ecosystems studied in this paper are within
the traditional range of the Inpuiaq and Yupik people. And, we would like to
acknowledge and thank them as the original stewards of these ecosystems. The
deployment of bio-logging devices used in this study were often done in
collaboration with Alaska Native seal hungers and their communities. We would
like to especially acknowledge the communities of Kotzebue, Koyuk, Utqiaġvik,
and Ulguniq (Wainwright) and the following individuals: James Adams, Jeff
Barger, David Barr, Wendell Booth, Cyrus Harris, Nereus 'Doc' Harris, Grover
Harris, Lee Harris, Tom Jones, Frank Garfield, Brenda Goodwin, Henry Goodwin,
John Goodwin, Pearl Goodwin, Willie Goodwin, Brett Kirk, Noah Naylor, Virgil
Naylor Jr., Virgil Naylor Sr., Dan Savetilik, Allen Stone, and Randy Toshavik
from Kotzebue, Alaska; Merlin Henry from Koyuk, Alaska; Billy Adams, Mary-Ellen
Ahmaogak, James Aiken, Tim Aiken, Sarah Coburn, Jason Herreman, Howard Kittick,
Gilbert Leavitt, Isaac Leavitt, J.R. Leavitt, Enoch Oktollik, Shawn Oktollik,
Stacey Osborn, Fred Rexford, Chuck Schaeffer, Ross Schaeffer, Bob Shears, and
Joe Skin from the North Slope Borough, Alaska;

We are grateful for the assistance in catching and sampling seals by
James Bailey, Michelle Barbieri, John Bengtson, Gavin Brady, Vladamir
Burkanov, Cynthia Christman, Shawn Dahle, Rob Delong, Stacy DiRocco, Deb
Fauquier, Shannon Fitzgerald, Kathy Frost, Scott Gende, Tracey
Goldstein, Jeff Harris, Markus Horning, John Jansen, Shawn Johnson,
Charles Littnan, Lloyd Lowry, Erin Moreland, Mark Nelson, Lorrie Rea, Gay Sheffield,
Brent Stewart, and Dave Withrow. We also appreciate the commitment to
science and safety by all officers and crew of the NOAA Ship Oscar
Dyson, the NOAA Ship MacArthur II, and the RV Thomas G. Thompson.

@@

The findings and conclusions in the paper are those of the author(s) and do not
necessarily represent the views of the National Marine Fisheries Service, NOAA.
Any use of trade, product, or firm names does not imply an endorsement by the
U.S. Government. Funding for this study was provided by the U.S. National
Oceanic and Atmospheric Administration. The field work was conducted under the
authority of Marine Mammal Protection Act Research Permits Nos. 782-1676,
782-1765, 15126, and 19309 issued by the National Marine Fisheries Service, and
Letters of Assurance of Compliance with Animal Welfare Act regulations, Nos.
A/NW 2010-3 and A/NW 2016-1 from the Alaska Fisheries Science Center/Northwest
Fisheries Science Center Institutional Animal Care and Use Committee (IACUC). ).
Funding to ADFG for tagging seals was provided by the Bureau of Ocean Energy
Management (No. M13PC0015) and the Office of Naval Research (No.
N00014-16-1-3019). ADFG and NSB field work was covered by Research Permits Nos.
358-1585, 358-1787, 15324, and 20466 and by ADF&G IACUC permits Nos. 06-16,
09-21, 2014-03, 2015-25, 2016-23, 0027-2017-27, 0027-2018-29, 0027-2019-041.
